---
title: "Cleaning and Wrangling data"
author: "Collin J. Closek"
date: "2023-04-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load packages
```{r load packages, warning=FALSE, message=FALSE}
library(readr)
library(dplyr)
library(tidyr)
```




# Introduction
About the data:
This data was downloaded from this [AOOS link](https://knb.ecoinformatics.org/view/df35b.304.2#df35b.302.1) on April 4, 2023.


## Read data
```{r read data, warning=FALSE, message=FALSE}
catch_original <- read_csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1")
```

## Explore data
```{r explore, eval=FALSE, echo=FALSE}
colnames(catch_original)

head(catch_original)

summary(catch_original)

sort(unique(catch_original$Year))

unique(catch_original$Region)
```


# Clean and Reshape data



## Selecting and Removing columns with `select()`

```{r select}

catch_data <- catch_original %>%
  select(Region, Year, Chinook, Sockeye, Coho, Pink, Chum)

head(catch_data)

```


## Check column types

```{r checking}
glimpse(catch_data)
```
Since Chinook is a Character, we need to change the column to integer or dbl using    `mutate`

Changing column content using `mutate()`
```{r changing data type}
catch_clean <- catch_data %>% 
  mutate(Chinook = as.numeric(Chinook))

head(catch_clean)

```
 
## Checking NA
```{r checking na}

i <- which(is.na(catch_clean$Chinook))

catch_data[i,]

```
 
## Fixing Chinook column
```{r}
catch_clean <- catch_data %>%
  mutate(Chinook = if_else(Chinook == "I",
                           "1", 
                           Chinook),
         Chinook = as.numeric(Chinook))
# This if_else statement is saying using the Chinook column, if a value in the Chinook column is equal to "I", then replace with "1" or if not, then replace with the same value in the Chinook column. Ultimately, the Chinook column was a character rather than numeric, so you change the column to numeric.

# Check
catch_clean[i,]
```


## Reshape data
From wide to long use `pivot_longer`
```{r reshape}
catch_long <- catch_clean %>% 
  pivot_longer(cols = -c(Region, Year),
               names_to = "species",
               values_to = "catch")

head(catch_long)
```

From long to wide use `pivot_wider()'
```{r reshape wide}

catch_wide <- catch_long %>% 
  pivot_wider(names_from = species,
              values_from = catch)

head(catch_wide)
```


## Renaming column
```{r}
catch_long <- catch_long %>% 
  rename(catch_thousands = catch)

colnames(catch_long)

```


## Add column using `mutate()`

```{r adding column}

catch_long <- catch_long %>% 
  mutate(catch = catch_thousands*1000) %>% 
  select(-catch_thousands)


head(catch_long)
```


## Split-Apply-Combine workflow
We are going to use the `group_by()` and `summarize()`
```{r}

mean_region <- catch_long %>% 
  group_by(Region) %>% 
  summarize(catch_mean = mean(catch))

head(mean_region)

mean_reg_yr <- catch_long %>% 
  group_by(Region, Year) %>% 
  summarize(catch_mean = mean(catch))

head(mean_reg_yr)

n_region <- catch_long %>% 
  group_by(Region) %>% 
  summarize(number_of_observations = n())

head(n_region)
```

## Filtering by rows with `filter()`
```{r filtering}

sse_catch <- catch_long %>% 
  filter(Region == "SSE")

head(sse_catch)

```

