---
title: "Intro to sf package for spatial data and making maps"
author: "Collin J. Closek"
date: "2023-04-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries 
```{r libraries}
library(readr)
library(sf)
library(ggplot2)
library(leaflet)
library(scales)
library(ggmap) #used for base maps to static maps
library(dplyr)
```

## Read in data 
Shape files from [kb](https://dev.nceas.ucsb.edu/view/urn:uuid:6f07cb25-a4a1-48e8-95cb-74f532f3ce2d)
```{r}
ak_regions <- sf::read_sf("data/ak_regions_simp.shp")
#you can press tab after read_sf to navigate available files in the data folder
```

## Explore data
```{r}
plot(ak_regions)
```
run "class(ak_regions)" in the console

```{r}
head(ak_regions)
```
```{r}
glimpse(ak_regions)
```

**Unprojected vs. projected**
Every sf object needs a coordinate reference system (or crs) defined in order to work with it correctly.

Some crs are unprojected or geographic coordinate system (gcs) are 3D.

2D space is represented as projected.

Both unprojected and projected use lat/lon, however the unprojected lat/lon are where the space is on the globe vs. the projected are rendered for 2D flat images.


```{r}
st_crs(ak_regions)
```
Shows that there's Geo CRS data (i.e., 3D), which isn't good for 2D flat maps

### Transform the geo crs to projected crs using the [EPSG](https://spatialreference.org/ref/epsg/?page=1) code 3338
```{r}
ak_regions_3338 <- ak_regions %>% 
  st_transform(crs = 3338)

st_crs(ak_regions_3338)
```

```{r map projected}
plot(ak_regions_3338)
```

## `sf` & the tidyverse
```{r}
colnames(ak_regions_3338)
```

```{r}
ak_regions_3338 %>%
    select(region)
```
the geometery column is what makes it a sf object - it's "sticky" because the file is an sf file and using the sf package, so it stays with any column we select.

```{r}
unique(ak_regions_3338$region)
```

```{r}
ak_regions_3338 %>%
    filter(region == "Yukon")
```

You can learn more about the tidyverse methods for sf objects using `?sf::tidyverse`

## Spatial joins

## Read in pop data
```{r}
pop <- read_csv("data/alaska_population.csv")
```
`View(pop)` and `class(pop)` help us to understand more about the data an that it's not an sf object, only a data.frame

```{r}
pop_4326 <- st_as_sf(pop,
                     coords = c("lng", "lat"),
                     crs = 4326,
                     remove = FALSE)
```
You must always keep the coords in the order of longitude, then latitude.

`st_as_sf` allows you to convert the data into an sf object

```{r join the populations within the regions}
pop_joined <- st_join(pop_4326, ak_regions_3338, join = st_within)
```
`?st_join` will allow you to see the many join options within the `st_join` pkg.

Need to rerun because the crs are not matching, so you need to use 4326: WGS84 (World Geodetic System 1984) as a default point of reference and then to 3338: Alaska Albers (projected CRS).

```{r}
pop_3338 <- st_transform(pop_4326, crs = 3338)

st_crs(pop_3338)
```

```{r}
pop_joined <- st_join(pop_3338, ak_regions_3338, join = st_within)
```

Transform to a data.frame to do `group_by()` and `summarize()`
```{r transform to a data frame}
pop_region <- pop_joined %>% 
  as.data.frame() %>% 
  group_by(region) %>% 
  summarize(total_pop = sum(population))

head(pop_region)
```

```{r add pop_region to the spatial data}

pop_region_3338 <- left_join(ak_regions_3338, pop_region, by = "region")
```
`View(pop_region_3338)` allows you to see that the two objects (pop_region & ak_regions_3338) are now combined

```{r}
plot(pop_region_3338["total_pop"])
```

```{r}
pop_mgmt_3338 <- pop_region_3338 %>% 
  group_by(mgmt_area) %>% 
  summarize(total_pop = sum(total_pop), do_union = FALSE)

head(pop_mgmt_3338)
```
`do_unioin` adds on the region boundaries (i.e., geometries)


```{r}
plot(pop_mgmt_3338["total_pop"])
```

Export to a shape file
```{r}
write_sf(pop_region_3338, "data/ak_regions_population.shp")
```


## Visualize with ggplot

```{r}
#add scalebar using `ggsn` package
ggplot(pop_region_3338) +
  geom_sf(aes(fill = total_pop)) +
  labs(fill = "Total Populaiton") +
  scale_fill_continuous(labels = comma,
                        low = "khaki",
                        high = "firebrick") +
  theme_light()
```
Correct for colorblind
```{r}
ggplot(pop_region_3338) +
  geom_sf(aes(fill = total_pop)) +
  labs(fill = "Total Populaiton") +
  scale_fill_viridis_c(labels = comma) +
  theme_light()
```

You could log transform the population
```{r}
ggplot(pop_region_3338) +
  geom_sf(aes(fill = log10(total_pop))) +
  labs(fill = "Total Populaiton (log10)") +
  scale_fill_viridis_c(labels = comma) +
  theme_light()
```

## Rivers spatial data
```{r}
rivers <- read_sf("data/ak_rivers_simp.shp")

st_crs(rivers)

#it's projected in alaska Albers, so let's change the `rivers` object name to `rivers_3338` to stay consistant with the above objects that are projected

rivers_3338 <- rivers

rm(rivers)
```


```{r}
ggplot() +
  geom_sf(pop_region_3338, mapping=aes(fill = log10 (total_pop))) +
  geom_sf(pop_3338, mapping=aes(), color = "red3", size = 0.5) +
  geom_sf(rivers_3338, mapping=aes(linewidth = StrOrder)) +
  scale_linewidth(range = c(0.05, 0.5), guide = "none") +
  scale_fill_viridis_c(labels = comma) +
  labs(fill = "Total Population", title = "Map of Alaska Populations")+
  theme(
    plot.title = element_text(hjust = 0.5)
  )
#you are having to add "mapping=aes()" because there is nothing in ggplot() above. geom_sf is run twice because we are layering different data ontop of one another
```


## Leaflet interactive map

Used infrom from "Proj4" in the spatial reference [EPSG:3338 page](https://spatialreference.org/ref/epsg/3338/)
```{r}
epsg3338 <- leaflet::leafletCRS(
    crsClass = "L.Proj.CRS",
    code = "EPSG:3338",
    proj4def =  "+proj=aea +lat_1=55 +lat_2=65 +lat_0=50 +lon_0=-154 +x_0=0 +y_0=0 +ellps=GRS80 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs",
    resolutions = 2 ^ (16:7)
)
```

Leaflet requires the projected (above) as well as the unprojected transformed from crs = 4326
```{r}
pop_region_4326 <- pop_region_3338 %>% st_transform(crs = 4326)
```

```{r}
m <- leaflet(options = leafletOptions(crs = epsg3338)) %>% 
  addPolygons(data = pop_region_4326,
              fillColor = "lightgreen",
              weight = 2,
              fillOpacity = 2) #%>% 
  #addTiles()

m

```



