---
title: 'R Practice: Spatial Data and Making Maps'
author: "Collin J. Closek"
date: "2023-04-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries
```{r libraries}
library(readr)
library(dplyr)
library(sf)
library(ggplot2)
library(scales)
library(leaflet)
library(ggmap)
```

## Bring in data
```{r}
pika <- read_csv("data/pika_fecal_glut.aw.data.csv")
elevation <- read_csv("data/pika_fecal_glut_habitat.aw.data.csv")

#alternatively, we could have brought in the linked data via:
#pika <- read_csv("https://portal.edirepository.org/nis/dataviewer?packageid=knb-lter-nwt.268.1&entityid=43270add3532c7f3716404576cfb3f2c")
#elevation <- read_csv("https://portal.edirepository.org/nis/dataviewer?packageid=knb-lter-nwt.268.1&entityid=6a10b35988119d0462837f9bfa31dd2f")

```

## Explore data
```{r explore}
head(pika)
glimpse(pika)
unique(pika$Site)
names(pika)


head(elevation)
glimpse(elevation)
unique(elevation$Station)

```

## Subset pika data
```{r}
pika_clean <- select(pika, c("Date", "Station", "UTM_Easting",        "UTM_Northing", "Concentration_pg_g", "Site", "Sex"))
head(pika_clean)
```

```{r}
glimpse(pika_clean)
```

## Rechlass Station, Site, Sex as `factor`, rather than `character`
```{r clean pika}
pika_clean <- pika_clean %>% 
  mutate(
    Station = as.factor(Station),
    Site = as.factor(Site),
    Sex = as.factor(Sex)
    )

glimpse(pika_clean)
```
## Recode Site locations with full name
```{r}
unique(pika_clean$Site)

pika_clean <- pika_clean %>% 
  mutate(
    Site = recode(
      Site,
      "CG" = "Cable Gate", 
      "WK" = "West Knoll", 
      "LL" = "Long Lake"
      ))

glimpse(pika_clean)
```

#Move `Station` & `Site` columns next to one another
```{r}
pika_clean %>% 
  relocate(Site, .after = Station)
```

## Reorder the columns
```{r}
pika_clean %>% 
  relocate(Site, .before = Station) %>% 
  relocate(Sex, .after = UTM_Northing)
```
## Subset `elevation` 
```{r}
elevation_clean <- elevation %>% 
  select("Station", "Elev_M")

glimpse(elevation_clean)
```

## Change `Station` from category `character` to `factor`
```{r}
elevation_clean <- elevation_clean %>% 
  mutate(
    Station = as.factor(Station)
  )

glimpse(elevation_clean)
```

# Join the two objects together via the column `Station` 
```{r}
pika_joined <- pika_clean %>% 
  full_join(y = elevation_clean,
            by = "Station")

#alternatively: 
#  pikas <- full_join(x = pika_clean,
#                   y = elevation_clean,
#                   by = "Station")
```

## Create Pikas Map
```{r}
class(pika_joined)

pikas_sf <- st_as_sf(x= pika_joined,
                     coords = c("UTM_Easting", "UTM_Northing"))

```
coords must be in the order of lon and then lat

## Set the CRS for the geometry column 
use the EPSG code 26913
```{r}
#st_crs(pikas_sf)

pikas_sf <- st_set_crs(pikas_sf, 26913)

ggplot(data = pikas_sf) +
    geom_sf()
```

## Plotting points
```{r}
ggplot(data = pikas_sf) +
    geom_sf(aes(color = Site, shape = Site), alpha = 0.6) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 30))
```


```{r}
ggplot(data = pikas_sf) +
    geom_sf(aes(color = Site, shape = Site), alpha = 0.6) +
    theme_minimal() +
    labs(
        title = "Location of Pika Sampling Stations",
        subtitle = "Niwot Ridge LTER",
        x = "Latitude (Degrees)",
        y = "Longitude (Degrees)"
    ) +
    theme(axis.text.x = element_text(angle = 30))
```

## Add a basemap using `ggmap`
```{r}
pikas_sf_4326 <- st_transform(pikas_sf, 4326)
nwt_bbox <- c(left = -105.65, bottom = 40.04, right = -105.55, top = 40.1)
```

```{r}
nwt_basemap <- ggmap::get_stamenmap(nwt_bbox, zoom = 13, 
                                    maptype = "terrain")

ggmap(nwt_basemap)
```

```{r}
ggmap(nwt_basemap) +
    geom_sf(pikas_sf_4326, 
            mapping = aes(color = Site, shape = Site), 
            inherit.aes = FALSE)
```

```{r}
ggmap(nwt_basemap) +
    geom_sf(pikas_sf_4326,
            mapping = aes(color = Site, shape = Site),
            inherit.aes = FALSE) +
    theme_minimal() +
    labs(
        title = "Location of Pika Sampling Stations",
        subtitle = "Niwot Ridge LTER",
        x = "Longitude (Degrees)",
        y = "Latitude (Degrees)"
    ) +
    scale_color_manual(values = c("black", "red", "purple")) 
```



