---
title: "RMarkdown Analysis"
author: "Collin J. Closek"
date: "2023-04-03"
output: 
  html_document:
    toc: true
    toc_float: true

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This contains North Pole Environmental Observatory Bottle Chemistry data which includes measurements of seawater chemistry along with supporting conductivity, temperature, and depth (CTD) data. Measurements were taken from sea ice platforms each April or May from 2000-2014. More information about the project is available at the [project web site link](http://psc.apl.washington.edu/northpole).

This [data set](https://arcticdata.io/catalog/view/doi:10.18739/A25T3FZ8X) was downloaded on April 3, 2023.

### Read data

#### Set up
```{r library}
library(readr)
```

```{r read data, warning=FALSE, message=FALSE}
bg_chem <- read_csv("data/BGchem2008data.csv")
```

#### Data exploration
```{r data expl, eval=FALSE, echo=FALSE}
#eval is whether to run (or not) and echo is whether to print the code in the RMrkdwn

## column names
colnames(bg_chem)

## structure of my data
str(bg_chem)

## summary stats
summary(bg_chem)

## Open data in new tab
View(bg_chem)

## unique values in a column
unique(bg_chem$Date)
```


## Analysis

### Calculate summary statistics
```{r}
nitrate <- mean(bg_chem$NO3)
nitrite <- mean(bg_chem$NO2)
amm <- mean(bg_chem$NH4)
phos <- mean(bg_chem$P)
```


### Calculate mean Redfield ratio
Redfield ratio is the total nitrogen over phosphorous in a sample (16:1)
```{r ratio calc}

ratio <- (amm + nitrate + nitrite)/phos

```


### Plot Redfield ratio
```{r plot}
plot(bg_chem$P, bg_chem$NO3 + bg_chem$NO2 + bg_chem$NH4)
```


## Conclusion

The Redfield ration is `r round(ratio)`